Nextstrain: real-time tracking of pathogen evolution

### 2018-04-17 replyToReviewer.tex

\documentclass[12pt,a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\begin{document}

We thank the reviewer for their kind words, and have addressed all of the points (detailed below).

\begin{quote}
1. Page 1, Line 39: ``ongoing and ever-present" is somewhat redundant; I'd just go with ``ever-present"

2. Page 1, Line 40: perhaps change ``recent epidemics" to ``recent events" so ``epidemics" isn't repeated 3x in the sentence.
\end{quote}
We have changed the wording of the manuscript to address these points.

@@

\begin{quote}
3. Page 1, Line 42: it might be helpful to point the reader to a general reference on phylodynamics, assuming the reference limit hasn't already been reached.
\end{quote}
We have reworded this section to include a reference to ``Viral Phylodynamics'', Voltz \textit{et al.} 2013.

\begin{quote}
4. Page 1, Line 43 (column 2): could references to the three databases (papers or URLs) be included here?
\end{quote}
We have added URLs to the text.

\begin{quote}
5. Page 2, Line 35: if it can be explained in a very short sentence, it might be interesting to mention the caveat of ``nearly all outbreaks" - which ones aren't amenable to Nextstrain-ification?
\end{quote}
There are indeed pathogens for which the general approach outlined in this paper is insufficient, and so we have changed the text to read ``Nextstrain is able to be extended to most outbreaks with readily accessible genomic data, although we note the potential for recombination or low mutation rate to confound phylogenetic signal.''

@@

\begin{quote}
6. Page 2, Line 51: it should be noted that the transmission inference depends on how well the outbreak in question was sampled - if data is missing, important transmission links could be missed.
\end{quote}
This is a very important point, and we have added the following sentence to the manuscript: ``It is important to note that sampling bias can obscure transmission links, and in certain cases we have chosen not to display the inferred states due to this bias.''

\begin{quote}
7. Page 2, Line 20 (column 20): is there an example of one of these mutations that could be mentioned here?
\end{quote}
We have added a reference to the 2017 publication ``Three mutations switch H7N9 influenza to human-type receptor specificity'' (de Vries \textit{et al.}), however have not listed these mutations as we detect none in the (consensus) sequences shown on nextstrain.org

### 2018-04-17 whitepaper.tex
insignificant changes in response

Nextstrain consists of data curation, analysis and visualisation components:
Python scripts maintain a database representative of all available sequences and related metadata, sourced from public repositories such as NCBI (www.ncbi.nlm.nih.gov), GISAID (www.gisaid.org) and ViPR (www.viprbrc.org), as well as GitHub repositories and other sources of genomic data.
A suite of tools perform phylodynamic analysis \citep{voltz2013}, including subsampling, alignment, phylogenetic inference, temporal dating of ancestral nodes and discrete trait geographic reconstruction, including inference of the most likely transmission events.
This leverages the maximum likelihood analyses implemented in TreeTime \citep{sagulenko2017treetime}, allowing a full analysis of the entire Ebola epidemic ($n$=1581 genomes) in under 2 hours on a modern laptop.
These scripts separate generic core functionality from a light pathogen-specific layer such that they are easily co-opted to different pathogens.
Visualisation of these analysis is available through \href{http://nextstrain.org}{nextstrain.org}.
This approach is similar in concept to Nextflu \citep{neher2015nextflu} however extended and generalised to different viral pathogens.
There is a growing need for surveillance of non-influenza viruses \citep{tang2017global}, and Nextstrain is able to be extended to most outbreaks with readily accessible genomic data, although we note the potential for recombination or slow mutation rate to confound phylogenetic signal.

>>>>

Nextstrain consists of data curation, analysis and visualisation components:
Python scripts maintain a database representative of all available sequences and related metadata, sourced from public repositories such as NCBI (www.ncbi.nlm.nih.gov), GISAID (www.gisaid.org) and ViPR (www.viprbrc.org), as well as GitHub repositories and other sources of genomic data.
A suite of tools perform phylodynamic analysis \citep{voltz2013}, including subsampling, alignment, phylogenetic inference, temporal dating of ancestral nodes and discrete trait geographic reconstruction, including inference of the most likely transmission events.
This leverages the maximum likelihood analyses implemented in TreeTime \citep{sagulenko2017treetime}, allowing a full analysis of the entire Ebola epidemic ($n$=1581 genomes) in under 2 hours on a modern laptop.
These scripts separate generic core functionality from a light pathogen-specific layer such that they are easily co-opted to different pathogens.
Visualisation of these analysis is available through \href{http://nextstrain.org}{nextstrain.org}.
This approach is similar in concept to Nextflu \citep{neher2015nextflu} however extended and generalised to different viral pathogens.
There is a growing need for surveillance of non-influenza viruses \citep{tang2017global}, and Nextstrain is able to be extended to most outbreaks with readily accessible genomic data, although we note the potential for recombination or low mutation rate to confound phylogenetic signal.
